Numerical modeling of artificial heart valve

### 2015-05-17 citech-2015/Paper-template/paper.tex
Translators corrections

% first the title is needed
\title{Numerical simulation of the artificial heart valve dynamics}

@@

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{D.A. Dolgov \and Y.N. Zakharov}
%
\authorrunning{CITech-2015, Almaty, Kazakhstan, 2015 September 24-27}
% (feature abused for this document to repeat the title also on left hand pages)

@@

\begin{abstract}
    In this paper we consider a mathematical model for describing the dynamics
    of the artificial heart valve, which is moved by the viscous inhomogeneous
    incompressible fluid flow with variable viscosity, and also a method
    of the numerical simulation of this problem. We present the results of modeling
    for the heart valve with three leaflets.
\keywords{viscous inhomogeneous fluid, artificial heart valve, immersed boundary method}
\end{abstract}

\section{Introduction}
It is difficult to overestimate the importance of medical researches of the human blood circulatory system,
because the knowledge from this area are extremely practical and significant. Each year, approximately 250 000
procedures are performed in the world to repair or replace damaged heart valves \cite{yoganathan}, and this
value tends to increase \cite{yacoub}. The solution of scientific and technical problems of the artificial valves 
creation depends on correct understanding of the fluid flow interaction with valve leaflets. Mathematical modeling
of the dynamics of artificial heart valves allows getting more complete understanding of processes inside them and helps
to find the ways to improve their design. There are many researches devoted to the mathematical and numerical modeling
of the dynamics of the heart valve. Most of them can be divided into two large groups.

First group is related to the finite element methods (\cite{taylor}, \cite{zhang}, \cite{black}). They may well handle
the complex geometry of the heart, bu the necessity to take into account the interaction between the fluid and flexible walls
leads to permanent rebuilding of the computational grid to meet the changing geometry of the object of research.
This results to significant expenses of time and computational resources.

In this paper we consider a second approach, which is related to the immersed boundary method (\cite{pescin_1977},
\cite{boyce_2011}, \cite{ma_x_2013}, \cite{pilhwa_2010}). It can be used for the problems with complex geometry, and it doesn't
require the grid modification.

There are various improvements of this method, because of the need to model more and more complex problems. In the research \cite{fai_2013}
a formulation of this method was proposed for the three dimensional problem of the flow of two non mixed (separated by flexible barriers)
fluids of different viscosity and density. In the papers \cite{jian}, \cite{lee} an application of this method for the two dimensional problem of
the two component fluid flow.

We propose to describe the blood flow in the flexible large blood vessels and the artificial heart valve as a three dimensional
nonstationary flow of the viscous incompressible fluid with variable viscosity and density (see \cite{gummel}, \cite{geidarov},
\cite{milosevic}, \cite{dolgov}). Thus, the goal of this work is to build a mathematical model and a solution method of the problem
of artificial heart leaflet dynamics inside a blood vessel in view of the inhomogeneous structure of the blood, and also about the
mixture (formed elements) motion inside vessel.

\section{Mathematical formulation of the problem}

We consider a nonstationary problem of blood flow inside vessel with valve. Blood consists of the plasma and formed elements, which are approximately
45\% of the entire volume \cite{caro}. Vessel walls and valve leaflets consists of the large number of thin collagen fibers, they are flexible
and can change their form depending on the fluid flow. The aortic valve, which are placed at the outlet from the left ventricle to the aorta, and provide
one-way movement of the blood, can be an example of this kind of systems (see Fig. \ref{fig:aortic_valve_example}):

@@

We model the blood as a viscous incompressible inhomogeneous two component fluid with variable viscosity, and vessel wall and valve leaflets as a
liquid impermeable surface with specified stiffness. Vessel and valve leaflets are deformed by a fluid pressure.

\begin{figure}
\centering
\includegraphics[width=12.5cm]{images/area_3d.png}
\caption{Scheme of the boundaries of the computational domain}
\label{fig:area_3d}
\end{figure}

Because the source of the blood motion in vessels is the pressure during the cardiac cycle, we will describe the problem of the blood flow
by the following nonstationary system of differential equations of Navier-Stokes \cite{gummel}:
\begin{gather}
    \label{eq:navier_stokes:motion}
    \frac{\partial \vec{u}}{\partial t} + (\vec{u} \cdot \nabla) \vec{u} = - \frac{1}{\rho} \nabla p + \nabla \sigma + \vec{f}\\
    \label{eq:navier_stokes:continuity}
    \frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \vec{u}) = 0 
\end{gather}

@@

where $\bar{x}=(x,y,z) \in \Omega$, $\vec{u}=(u,v,w)$ - velocity vector, $\vec{u}_b$ - velocity of the vessel wall and valve leaflet motion during the deformation,
$\rho=\rho(\bar{x}, t)$ - density, $p=p(\bar{x}, t)$ - pressure, $\sigma = \mu (\nabla \vec{u} + (\nabla \vec{u})^T)$ - viscous stress tensor,
$\mu = \mu(\bar{x}, t)$ - viscosity of fluid, $\vec{f} = \vec{f}(\bar{x}, t)$ - vector of body forces, which is further used to determine form of the vessel and valve leaflets.
Domain $\Omega$ is a vessel with boundary $\Gamma = \Gamma_1 \cup \Gamma_2 \cup \Gamma_3 \cup \Gamma_4$, where $\Gamma_1$ - vessel wall, $\Gamma_2$ and $\Gamma_3$ - inlet and outlet, $\Gamma_4$ - valve leaflets (see Fig. \ref{fig:area_3d}).
As shown in \cite{ragulin}, the problem (\ref{eq:navier_stokes:motion}) - (\ref{eq:navier_stokes:continuity}) has a unique solution.

@@

where $\rho_1$, $\mu_1$ - density and viscosity if fluid (plasma), $\rho_2$, $\mu_2$ - density and viscosity of mixture (formed elements), $c$ - concentration of mixture. Concentration $c=c(\bar{x}, t)$, $c \in [0, 1]$ of mixture are determined as a solution of equation:
\begin{gather}
    \label{eq:convection}
    \frac{\partial c}{\partial t} + \vec{u} \cdot \nabla c = 0
\end{gather}

@@

and boundary conditions at the inlet boundary:
\begin{gather}
    \label{eq:convection:conditions}
    c(\bar{x}, t)|_{\Gamma_2} = c_s(\bar{x}, t)
\end{gather}

where $c_0, c_s$ are predefined functions.

Lack of definition for one component of velocity vector at the inlet-outlet is the one of issues for computational solution of this kind of problems.
It can be solved by the using the original equations (\ref{eq:navier_stokes:motion}) - (\ref{eq:navier_stokes:pressure_conditions}) at the boundaries
$\Gamma_2$, $\Gamma_3$ for determine of missing components of the velocity vector (see details \cite{gummel}).

Motion of the vessel walls and valve leaflets is defined by forces, which returned it to the original position. Valve leaflets can be deformated much more,
than vessel walls. To describe the forces, arising due the valve deformation, we use following formula:

@@

where $\bar{q} = (q, r, s) \in \Gamma_4$, $X(\bar{q})$ - function for describing the surface of leaflets at the moment $t$, the coordinates $q, r, s$ chosen so
that the surface $X$ was presented by the large amount of parametric lines $s \rightarrow X(q^0, r^0, s)$, $T$ - tension, that arises due the stretching along $s$,
$E$ - Young's modulus, $I$ - cross-sectional moment of inertia (see \cite{boyce_2011}, \cite{pescin_2002}). Physically the formula above means,
that the valve leaflets are resist extension, compression (it's related to the first term with $T$, which is dependent on stiffness coefficient $k$),
and bending (this related to the second term, where $E$ and $I$ are referred as a stiffness coefficient $k_b$). Formula (\ref{eq:boundary_force}) allows taking
into account any changes of the valve shape.

To compute forces, arising due the deformation of the vessel, we use another formula, which allows taking into account only small changes of the shape:

@@

As shown in the works \cite{pescin_1977}, \cite{boyce_2011}, for modeling of the interaction between vessel walls, valve leaflets and the fluid flow we need
to compute the vector of body forces $f$ in the equation of Navier-Stokes, based on the force $F$, and determine current surface $X(\bar{q}, t)$ of the vessel and valve, based on the velocity vector field $\vec{u}(\bar{x}, t)$. This is done by using the following equations:
\begin{gather}
    \label{eq:interaction:velocity}
    \frac{\partial X}{\partial t}(\bar{q}, t) = \int_{\Omega} \vec{u}(\bar{x}, t) \cdot \delta (x - X(\bar{q}, t))\; dx\; dy\; dz\\
    \label{eq:interaction:force}
    \vec{f}(\bar{x}, t) = \int_{\Gamma} \vec{F}(\bar{q}, t) \cdot \delta (x - X(\bar{q}, t))\; dq\; dr\; ds
\end{gather}

where $\bar{q} = (q, r, s) \in \Gamma$ - point at the vessel wall or valve leaflet, $X = X(\bar{q}, t)$ - function for describing the surface
of vessel wall and valve leaflet at the momen $t$, $F = F(\bar{q}, t)$ - the force of resistance to deformation,
$\vec{u}(\bar{x}, t)$ - fluid velocity vector, $\vec{f}(\bar{x}, t)$ - vector of body forces, $\delta$ - Dirac delta function.

Thus, we build the model, which describes the motion of the viscous inhomogeneous incompressible fluid inside vessel with valve. In this model the fluid state
and the surface $\Gamma_1 \cup \Gamma_4$ are determined independently from each other, and the influence of the valve leaflets on the fluid is reflected by
relation (\ref{eq:interaction:force}) between the vector of body forces $\vec{f}(\bar{x}, t)$ from the equation (\ref{eq:navier_stokes:motion}) and the force of resistance to deformation $F = F(\bar{q}, t)$ from the equations (\ref{eq:boundary_force}), (\ref{eq:boundary_force_simple}).

\section{Method of solution}

As it was mentioned before, in this work we using the immersed boundary method \cite{pescin_1977}, which are based on that when fluid flows over a body,
it exerts a normal force on the surface, and if the surface is no-slip, the fluid also exerts a shear force. The surface exerts the same force of opposite sign.
This means that fluid flow around a body can be modeled by a corresponding field of the external body forces \cite{goldstain}.

Accordingly to the immersed boundary method, we will determine the fluid flow in the parallelepiped $\tilde{\Omega}$, which contains $\Omega$.
No-slip conditions are imposed at the boundaries if $\tilde{\Omega}$.
For the computation of the fluid flow we will use rectangular uniform staggered grid $\tilde{\Omega_h}$ with grid spacing $h_x$, $h_y$, $h_z$ and 
staggered arrangement of cells, where the pressure, velocity divergence and concentration are computed at the center of cell, the velocity vector components
and vector of external forces - at the boundaries of cell. To determine the deformation of the surface $\Gamma_1 \cup \Gamma_4$ we will introduce additional
area $\tilde{\Gamma}$ with Lagrangian coordinate system, which is related to the vessel walls and valve leaflets. In the $\tilde{\Gamma}$ we will construct
a new grid $\tilde{\Gamma_h}$, which cells are corresponding to the points at the $\Gamma_1 \cup \Gamma_2$. Algorithm of solution consists of the several steps:
at the grid $\tilde{\Gamma_h}$ we will solve the problem (\ref{eq:navier_stokes:motion})-(\ref{eq:navier_stokes:pressure_conditions}); then we will solve
the convection equation (\ref{eq:convection}), i.e. determine the concentration of mixture and recalculate the density and viscosity. After it we will use
the formulas (\ref{eq:boundary_force}), (\ref{eq:boundary_force_simple}) and (\ref{eq:interaction:velocity}), (\ref{eq:interaction:force}) to determine position of
leaflets and the vessel form.

Differential equation (\ref{eq:boundary_force}), (\ref{eq:convection:conditions}) is solved by the finite difference method.
To solve (\ref{eq:boundary_force}), (\ref{eq:boundary_force_simple}) we will use the splitting on physical factors scheme \cite{belotserkovsky}:
\begin{gather}
    \label{eq:splitting:intermediate_velocity}
    \frac{u^* - u^n}{\triangle t} = - (u^n \cdot \nabla) u^* - \frac{1}{\rho} \nabla \sigma + f^n\\
    \label{eq:splitting:poisson}
    \rho \triangle p^{n+1} - \nabla \rho \cdot p^{n+1} = \frac{\rho^2 \nabla u^*}{\triangle t}\\
    \label{eq:splitting:velocity}
    \frac{u^{n+1} - u^*}{\triangle t} = - \frac{1}{\rho} \triangle p^{n+1}
\end{gather}

Computational implementation of this scheme consists of three stages. At the beginning the intermediate field $u^*$  is computed from the known values of velocity
from the previous time step. For this equation (\ref{eq:splitting:intermediate_velocity}) is solved by the method of stabilizing corrections \cite{yanenko}.
After it a new pressure field is determined via the computational solution of (\ref{eq:splitting:poisson}) with the biconjugate gradient method usage.
And at the last stage a final velocity vector field is calculated by the formula (\ref{eq:splitting:velocity}).

After fluid flow parameters determining it is necessary to calculate new values of density and velocity. To do that a new time step for the convection
equation (\ref{eq:convection}) must be done using the obtained values of velocity components, and the density and viscosity are recalculated
by the formulas (\ref{eq:viscosity}), (\ref{eq:density}).

Next we need to determine the deformation of vessel walls and valve leaflets under influence of fluid flow, and also the distribution of body forces $f$
in the equation of fluid motion based on this deformation. We can calculate the deformation of vessel walls and valve leaflets at this particular fluid pressure
and the appearing resistance forces using the equations (\ref{eq:interaction:velocity}) - (\ref{eq:interaction:force}), which are numerically integrated by
the any of quadrature formulas, and equations (\ref{eq:boundary_force}) - (\ref{eq:boundary_force_simple}). After it we recalculate the body forces $f$
and go to the next time step.

@@

We present several results of methodical calculations for the cases of constant and variable density and viscosity, whose purpose is to demonstrate the operability
of the described method and the possibility to get with its help the patterns of  leaflet deformation and mixture distribution inside the valve.
All calculations were performed in dimensionless variables. As a vessel, in which the valve was located, a circular cylinder with length $l = 1$,
radius $r = 0.11$ and wall stiffness $k = 1 \cdot 10^3$ was used, the domain $\tilde{\Omega}$ had the spatial parameters $1.0 \times 0.5 \times 0.5$,
spatial steps $h_x = h_y = h_k = 0.01$, time step $\triangle t = 0.01$.

The dynamics of valve with three leaflets under the influence of the fluid pressure with constant density and velocity is shown in Fig. \ref{fig:valve} and Fig. \ref{fig:valve_with_particles}.
The pressure differential $p_{in} - p_{out}$ changes periodically from 0 to 6. Coefficient of stretching resistance $k_b = 5 \cdot 10^3$ and coefficient of
bending resistance $k_b = 5 \cdot 10^3$ are specified for the valve leaflets.

@@

\caption{Dynamics of the Valve leaflets. Current leaflet shape is indicated by points, flow direction - by arrows.
It shows a side view (I), front view (II) and rear view (III). $k_s = 5 \cdot 10^3$, $k_b = 5 \cdot 10^3$, $\rho_1 = \rho_2 = 1$,
$\mu_1 = \mu_2 = 1 \cdot 10^{-2}$; a) $t=0$, b) $t=0.7$, c) $t=1.5$}
\label{fig:valve}
\end{figure}

@@

\caption{Tracks of particles inside the valve. Flow direction is indicated by arrows. Calculation parameters are the same as for the Fig. \ref{fig:valve}
It shows a side view (I) and front view (II). $k_s = 5 \cdot 10^3$, $k_b = 5 \cdot 10^3$, $\rho_1 = \rho_2 = 1$,
$\mu_1 = \mu_2 = 1 \cdot 10^{-2}$; a) $t=0$, b) $t=0.7$, c) $t=1.5$}

@@

As can be seen from the Fig. \ref{fig:valve} and Fig. \ref{fig:valve_with_particles}, the valve opens when pressure differential is increased,
and then returns to the original state when pressure is equalized.

The Fig. \ref{fig:valve_in_mixture} shows the dynamics of the valve with three leaflets under the fluid pressure with variable viscosity and
density. Pressure differential $p_{in} - p_{out}$ changes cyclically from 0 to 6. Coefficient of stretching resistance $k_s = 8 \cdot 10^3$ and coefficient of
bending resistance $k_s = 6 \cdot 10^3$ are specified for the valve leaflets. At the $\Gamma_2$ a constant mixture flow is imposed with concentration $c_s = 0.45$.

@@

\caption{Valve leaflets motion with variable viscosity and density. A constant mixture flow $c_s|_{\Gamma_2} = 0.45$ is imposed at the inlet,
concentration of mixture at the initial moment $c_0 = 0.45$, $\rho_1=1$, $\rho_2=1.2$, $\mu_1 = 1 \cdot 10^2$, $\mu_2 = 1.2 \cdot 10^2$;
a) $t = 4$, b) $t=5$, c) $t=6$}
\label{fig:valve_in_mixture}
\end{figure}

As you can see from the Fig. \ref{fig:valve_in_mixture}, initial uniform distribution of the mixture is disturbed by the leaflets motion. Over time
mixture motion begins to have an oscillatory character corresponding to the cycles of valve dynamics. Besides that it can bee seen,
that the mixture distribution along the section parallel to the axis $Oy$ is not symmetric, because the valve leaflets also are not symmetric about the axis $Oy$.

@@

Constructed model of blood flow with variable viscosity and density allows getting the patterns of leaflet deformation and mixture distribution 
under the influence of flow inhomogeneous fluid.

>>>>

% first the title is needed
\title{Numerical modeling of artificial heart valve}

@@

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{Dmitriy Dolgov \and Yury Zakharov}
%
\authorrunning{CITech-2015, Almaty, Kazakhstan, 2015 September 24-27}
% (feature abused for this document to repeat the title also on left hand pages)

@@

\begin{abstract}
    The paper is dedicated to the mathematical model describing dynamics of
    an artificial heart valve being moved by inhomogeneous incompressible fluid flow
    with variable viscosity, and its computational method. The modeling results of
    tricuspid valve performance are presented.
\keywords{viscous inhomogeneous fluid, artificial heart valve, immersed boundary method}
\end{abstract}

\section{Introduction}
The importance of medical researches of human blood circulatory system can hardly be overestimated,
because this kind of knowledge is extremely practical and significant. Annualy approximately 250 000
surgeries are performed in the world to restore or replace damaged heart valves \cite{yoganathan}, and the
quantity is expected to increase \cite{yacoub}. The solution of scientific and technical problems of the artificial valves 
creation depends on correct understanding of the interaction between blood flow and valve leaflets. Mathematical modeling
of artificial heart valves performance enables to get thorough understanding of its internal processes in order to improve its design.
There are many researches devoted to the mathematical and numerical modeling
of heart valve performance, based on which two main problem solution approaches were defined.

First approach is related to the finite element methods (\cite{taylor}, \cite{zhang}, \cite{black}). They enable to take into consideration
the complex geometry of heart, bu the necessity to take into account the interaction between fluid and flexible walls
requires constant rebuilding of the computational grid to meet the changing geometry of the object of research.
It appears to be time and computational resources consuming.

(The second approach, which is related to the immersed boundary method, is under discussion in this paper (\cite{pescin_1977},
\cite{boyce_2011}, \cite{ma_x_2013}, \cite{pilhwa_2010}). It can be used for the problems with complex geometry, and it doesn't
require grid modification.

There are various improvements of this method, in order to model more and more complex problems. In the research \cite{fai_2013}
a formulation of this method was proposed for the three dimensional flow problem of two non mixed (separated by flexible barriers)
fluids of different viscosity and density. In the papers \cite{jian}, \cite{lee} this method application in case of the two dimensional 
problem of two component fluid flow is presented.

We propose to describe the blood flow in the flexible large blood vessels and the artificial heart valve as a three dimensional
nonstationary flow of viscous incompressible fluid with variable viscosity and density (see \cite{gummel}, \cite{geidarov},
\cite{milosevic}, \cite{dolgov}). Thus, the goal of this work is to develop a mathematical model and a solution method of the problem
of artificial heart leaflet dynamics inside a blood vessel taking into account the inhomogeneous structure of the blood, and 
admixture (formed elements) circulation inside a blood vessel.

\section{Formulation of the problem}

We consider a nonstationary problem of blood flow inside a vessel with a valve. Blood consists of plasma and formed elements, which are approximately
45\% of the entire volume \cite{caro}. Vessel walls and valve leaflets consist of a large number of thin collagen fibers, they are flexible
and can change their form depending on the fluid flow. For example, the tricuspid aortic valve, lies between the left ventricle and the aorta, and 
prevents blood backflow (see Fig. \ref{fig:aortic_valve_example}):

@@

We model the blood as a viscous incompressible inhomogeneous two component fluid with variable viscosity, and vessel wall and valve leaflets as a
fluid impermeable surface with specified stiffness. Vessel and valve leaflets are deformed under the fluid pressure.

\begin{figure}
\centering
\includegraphics[width=12.5cm]{images/area_3d.png}
\caption{Computational domain boundaries}
\label{fig:area_3d}
\end{figure}

Since blood circulates through vessels under the pressure created by cardiac beats then the problem of the blood flow
can be described by Navier-Stokes nonstationary system of differential equations \cite{gummel}:
\begin{gather}
    \label{eq:navier_stokes:motion}
    \frac{\partial \vec{u}}{\partial t} + (\vec{u} \cdot \nabla) \vec{u} = - \frac{1}{\rho} \nabla p + \nabla \sigma + \vec{f}\\
    \label{eq:navier_stokes:continuity}
    \frac{\partial \rho}{\partial t} + \nabla \cdot (\rho \vec{u}) = 0 
\end{gather}

@@

where $\bar{x}=(x,y,z) \in \Omega$, $\vec{u}=(u,v,w)$ - velocity vector, $\vec{u}_b$ - velocity of the vessel walls and valve leaflets motion under deformation,
$\rho=\rho(\bar{x}, t)$ - density, $p=p(\bar{x}, t)$ - pressure, $\sigma = \mu (\nabla \vec{u} + (\nabla \vec{u})^T)$ - viscous stress tensor,
$\mu = \mu(\bar{x}, t)$ - fluid viscosity, $\vec{f} = \vec{f}(\bar{x}, t)$ - body forces vector, which is further used to determine form of the vessel and valve leaflets.
Domain $\Omega$ is a vessel with boundary $\Gamma = \Gamma_1 \cup \Gamma_2 \cup \Gamma_3 \cup \Gamma_4$, where $\Gamma_1$ - blood vessel wall,
$\Gamma_2$ and $\Gamma_3$ - inflow and outflow domains, $\Gamma_4$ - valve leaflets (see Fig. \ref{fig:area_3d}).
As shown in \cite{ragulin}, the problem (\ref{eq:navier_stokes:motion}) - (\ref{eq:navier_stokes:continuity}) has a unique solution.

@@

where $\rho_1$, $\mu_1$ - fluid density and viscosity (plasma), $\rho_2$, $\mu_2$ - admixture density and viscosity (formed elements), $c$ - admixture concentration.
Admixture concentration $c=c(\bar{x}, t)$, $c \in [0, 1]$ is determined as a solution of equation:
\begin{gather}
    \label{eq:convection}
    \frac{\partial c}{\partial t} + \vec{u} \cdot \nabla c = 0
\end{gather}

@@

and boundary conditions at the inflow boundary:
\begin{gather}
    \label{eq:convection:conditions}
    c(\bar{x}, t)|_{\Gamma_2} = c_s(\bar{x}, t)
\end{gather}

where $c_0, c_s$ are specified functions.

One of the issues determined for this kind of problem computational solutions is the lack of one component of velocity vector in the inflow-outflow areas.
It can be solved by using the original equations (\ref{eq:navier_stokes:motion}) - (\ref{eq:navier_stokes:pressure_conditions}) at the boundaries
$\Gamma_2$, $\Gamma_3$ to determine the missing components of the velocity vector (see details \cite{gummel}).

Motion of the vessel walls and valve leaflets is defined by the forces, which return them to the original position. Valve leaflets can be deformated much more,
than vessel walls. To describe the forces, arising due to the valve deformation, the following formula is used:

@@

where $\bar{q} = (q, r, s) \in \Gamma_4$, $X(\bar{q})$ - function for describing the valve leaflets surface at the moment $t$, the coordinates $q, r, s$ are chosen 
so that the surface $X$ is presented by the large amount of parametric lines $s \rightarrow X(q^0, r^0, s)$, $T$ - tension,
that arises due to the stretching along $s$, $E$ - Young's modulus, $I$ - cross-sectional moment of inertia (see \cite{boyce_2011}, \cite{pescin_2002}).
Physically the formula above means, that the valve leaflets resist stretching, (it's related to the first term with $T$,
which is dependent on stiffness coefficient $k$), and bending (it's related to the second term, where $E$ and $I$ are referred as a stiffness coefficient $k_b$).
Formula (\ref{eq:boundary_force}) allows taking into account any changes of the valve shape.

To compute the forces, arising due to the deformation of the vessel, another formula is used, which allows taking into account only small shape changes:

@@

Researches \cite{pescin_1977}, \cite{boyce_2011} show, that in order to the interaction between vessel walls, valve leaflets and the fluid flow it is necessary
to compute the field of external body forces $f$ in the Navier-Stokes equation, based on the force $F$, and determine the current form $X(\bar{q}, t)$ 
of the vessel and the valve, based on the fluid field of velocities $\vec{u}(\bar{x}, t)$. The following equations are used for this purpose:
\begin{gather}
    \label{eq:interaction:velocity}
    \frac{\partial X}{\partial t}(\bar{q}, t) = \int_{\Omega} \vec{u}(\bar{x}, t) \cdot \delta (x - X(\bar{q}, t))\; dx\; dy\; dz\\
    \label{eq:interaction:force}
    \vec{f}(\bar{x}, t) = \int_{\Gamma} \vec{F}(\bar{q}, t) \cdot \delta (x - X(\bar{q}, t))\; dq\; dr\; ds
\end{gather}

where $\bar{q} = (q, r, s) \in \Gamma$ - point at the vessel wall or valve leaflet, $X = X(\bar{q}, t)$ - function for describing the vessel
and valve surfaces at the moment $t$, $F = F(\bar{q}, t)$ - the force of deformation resistance at given point,
$\vec{u}(\bar{x}, t)$ - fluid flow velocity vector, $\vec{f}(\bar{x}, t)$ - body forces vector, $\delta$ - Dirac delta function.

Thus the model describing the motion of the viscous inhomogeneous incompressible fluid inside vessel with valve is built. This model enables to determine
the fluid state and the surface form $\Gamma_1 \cup \Gamma_4$ independently of each other. The valve leaflets influence on the fluid is described by 
correlation (\ref{eq:interaction:force}) between the vector of body forces $\vec{f}(\bar{x}, t)$ from (\ref{eq:navier_stokes:motion}) 
and the force of deformation resistance $F = F(\bar{q}, t)$ from (\ref{eq:boundary_force}), (\ref{eq:boundary_force_simple}).

\section{Solution method}

As it was mentioned before, immersed boundary method is used in the paper \cite{pescin_1977}. This method is based on the fact, that in case of flowing 
over a body the fluid is effected by surface force and shear force if the body has no-slip boundary condition. The body surface is influenced by the same
forces of opposite sign. It means that fluid flowing over the body can be modeled by a corresponding field of the external body forces \cite{goldstain}.

According to the immersed boundary method, we determine the fluid flow in the parallelepiped $\tilde{\Omega}$, which contains $\Omega$.
$\tilde{\Omega}$ has no-slip boundary conditions.
To compute of the fluid flow we use rectangular uniform staggered grid $\tilde{\Omega_h}$ with grid spacing $h_x$, $h_y$, $h_z$ and 
staggered arrangement of cells, where the pressure, velocity divergence and concentration are computed at the center of cell, the velocity vector components
and vector of external forces are computed at the boundaries of cell. To determine the deformation of the surface
$\Gamma_1 \cup \Gamma_4$ we introduce additional area $\tilde{\Gamma}$ with Lagrangian coordinate system, which is related to the vessel
walls and valve leaflets. In the $\tilde{\Gamma}$ we construct a new grid $\tilde{\Gamma_h}$, with cells corresponding to the points at the $\Gamma_1 \cup \Gamma_2$.
Solution algorithm consists of several steps:
at the grid $\tilde{\Gamma_h}$ the problem (\ref{eq:navier_stokes:motion})-(\ref{eq:navier_stokes:pressure_conditions}) is solved; then the convection equation 
(\ref{eq:convection}) is solved, i.e. the concentration of admixture is determined in the solution domain and the density and viscosity are recalculated.
Then formulas (\ref{eq:boundary_force}), (\ref{eq:boundary_force_simple}) and (\ref{eq:interaction:velocity}), (\ref{eq:interaction:force}) are used to
determine the position of leaflets and the vessel form.

Differential equation (\ref{eq:navier_stokes:motion}), (\ref{eq:convection:conditions}) is solved by the finite difference method.
To solve (\ref{eq:navier_stokes:motion}), (\ref{eq:navier_stokes:pressure_conditions}) splitting schemes due to physical factors are used \cite{belotserkovsky}:
\begin{gather}
    \label{eq:splitting:intermediate_velocity}
    \frac{u^* - u^n}{\triangle t} = - (u^n \cdot \nabla) u^* - \frac{1}{\rho} \nabla \sigma + f^n\\
    \label{eq:splitting:poisson}
    \rho \triangle p^{n+1} - \nabla \rho \cdot p^{n+1} = \frac{\rho^2 \nabla u^*}{\triangle t}\\
    \label{eq:splitting:velocity}
    \frac{u^{n+1} - u^*}{\triangle t} = - \frac{1}{\rho} \triangle p^{n+1}
\end{gather}

Numerical implementation of this scheme consists of three stages. At the beginning the intermediate field $u^*$  is computed using the known values of velocity
from the previous time step. Thus equation (\ref{eq:splitting:intermediate_velocity}) is solved by the method of stabilizing corrections \cite{yanenko}.
Then a new pressure field is determined via the computational solution of (\ref{eq:splitting:poisson}) using biconjugate gradient method.
At the last stage a final velocity vector field is calculated according to the formula (\ref{eq:splitting:velocity}).

As soon as fluid flow parameters are determined it is necessary to calculate new values of density and velocity. To do that a new time step for the convection
equation (\ref{eq:convection}) must be done using the obtained values of velocity components, and the density and viscosity are recalculated
by the formulas (\ref{eq:viscosity}), (\ref{eq:density}).

Next it is necessary to determine the deformation of vessel walls and valve leaflets being influenced by of fluid flow, and also the distribution of body forces $f$
in the fluid motion equation based on this deformation. It is possible to calculate the deformation of vessel walls and valve leaflets under this particular fluid pressure
and the resistance forces by using the equations (\ref{eq:interaction:velocity}) - (\ref{eq:interaction:force}), which are numerically integrated by
using any of quadrature formulas, and equations (\ref{eq:boundary_force}) - (\ref{eq:boundary_force_simple}). Afterwards, the body forces $f$
are recalculated, and move to the next time step.

@@

Some results of methodical calculations for the cases with constant and variable density and viscosity, which are aimed to demonstrate the described method 
validation and the possibility to get the patterns of leaflet deformation and admixture distribution inside the valve.
All calculations were performed in dimensionless variables. 
A circular cylinder with length $l = 1$, radius $r = 0.11$ and wall stiffness $k = 1 \cdot 10^3$ was used as a vessel with a valve,
the domain $\tilde{\Omega}$ had spatial parameters $1.0 \times 0.5 \times 0.5$,
spatial steps $h_x = h_y = h_k = 0.01$, time step $\triangle t = 0.01$.

Fig. \ref{fig:valve} and Fig. \ref{fig:valve_with_particles} show tricuspid valve dynamics effected by the pressure of fluid with constant density and velocity.
The pressure differential $p_{in} - p_{out}$ changes periodically from 0 to 6. Coefficient of stretching resistance $k_b = 5 \cdot 10^3$ and coefficient of
bending resistance $k_b = 5 \cdot 10^3$ are specified for the valve leaflets.

@@

\caption{Dynamics of the Valve leaflets. Current leaflet shape is indicated by points, arrows indicate flow direction.
Side view (I), front view (II) and rear view (III). $k_s = 5 \cdot 10^3$, $k_b = 5 \cdot 10^3$, $\rho_1 = \rho_2 = 1$,
$\mu_1 = \mu_2 = 1 \cdot 10^{-2}$; a) $t=0$, b) $t=0.7$, c) $t=1.5$}
\label{fig:valve}
\end{figure}

@@

\caption{Tracks of particles inside the valve. Flow direction is indicated by arrows. Calculation parameters are the same as in the Fig. \ref{fig:valve}
Side view (I) and front view (II). $k_s = 5 \cdot 10^3$, $k_b = 5 \cdot 10^3$, $\rho_1 = \rho_2 = 1$,
$\mu_1 = \mu_2 = 1 \cdot 10^{-2}$; a) $t=0$, b) $t=0.7$, c) $t=1.5$}

@@

As can be seen in the Fig. \ref{fig:valve} and Fig. \ref{fig:valve_with_particles}, the valve opens when pressure differential is increased,
and then reverts to the original state when pressure is balanced.

The Fig. \ref{fig:valve_in_mixture} shows the motion dynamics of the tricuspid valve under the pressure of fluid with variable viscosity and
density. Pressure differential $p_{in} - p_{out}$ changes cyclically from 0 to 6. Coefficient of stretching resistance $k_s = 8 \cdot 10^3$ and coefficient of
bending resistance $k_s = 6 \cdot 10^3$ are specified for the valve leaflets. Constant admixture flow with concentration $c_s = 0.45$ is set at $\Gamma_2$.

@@

\caption{Valve leaflets motion in a vessel with variable viscosity and density. A constant admixture flow $c_s|_{\Gamma_2} = 0.45$ at the inflow,
admixture concentration at the initial time $c_0 = 0.45$, $\rho_1=1$, $\rho_2=1.2$, $\mu_1 = 1 \cdot 10^2$, $\mu_2 = 1.2 \cdot 10^2$;
a) $t = 4$, b) $t=5$, c) $t=6$}
\label{fig:valve_in_mixture}
\end{figure}

Fig. \ref{fig:valve_in_mixture} shows, that initial uniform admixture distribution is interrupted by the valve leaflets motion. Eventually oscillatory mode of
admixture motion can be recognized that corresponds to valve operating cycle. 
Moreover the Fig. \ref{fig:valve_in_mixture} shows that admixture distribution over a cross section being parallel to axis $Oy$
is not symmetric because the valve leaflets are not symmetric about the axis $Oy$ as well.

@@

Constructed model of blood flow with variable viscosity and density allows to get the patterns of leaflet deformation and admixture distribution 
effected by inhomogeneous fluid flow.
